---
title: "Barplots"
output:
  pdf_document: default
  html_notebook: default
---

Code Notebook Author: Sara D. Williams

Contact: swilliams@mote.org

### Get your packages loaded in:
```{r}
library(palmerpenguins)
library(ggplot2)
```

### Inspect your data
```{r}
str(penguins)
```

### First, R's standard error function is a bit wonky, so make you're own.
```{r}
my.se<-function(x){
  sd(x,na.rm=TRUE)/sqrt(length(x))
}
```

### Barplot of penguin bill length (mm) mean and CI by species using baseplot
```{r}

sps.avg<-aggregate(bill_length_mm~species,data=penguins,FUN=mean) #get means for each species
sps.avg
sps.se<-aggregate(bill_length_mm~species,data=penguins,FUN=my.se) #get standard error

#par(family="Times New Roman") #Make the text pretty
bp<-barplot(as.matrix(t(sps.avg$bill_length_mm)), #need to transpose table
            ylim=c(0,50), #set for your dataset
            las=1, #changes angle of xlab text, = 2 for vertical
            ylab="Bill length (mm)", #change for your label, always include units
            names.arg=c(as.character(sps.avg$species)), #names to use for bars
            col="grey" )#bar colors
#add error bars for confidence interval, take out the 1.96* if just want SE
arrows(x0=bp,x1=bp,y0=(sps.avg$bill_length_mm)-1.96*(sps.se$bill_length_mm),
       y1=(sps.avg$bill_length_mm)+1.96*(sps.se$bill_length_mm),
       code = 3, angle = 90, len = 0.02, xpd = NA)

```

### Now barplot for species and sex
```{r}

sps.avg<-aggregate(bill_length_mm~species+sex,data=penguins,FUN=mean) #get means for each species
sps.avg
sps.sex.avg<-cbind(sps.avg[1:3,3],sps.avg[4:6,3])
sps.sex.avg
sps.se<-aggregate(bill_length_mm~species+sex,data=penguins,FUN=my.se) #get standard error
#can change these to use se instead
ci.low<-(sps.avg$bill_length_mm)-1.96*(sps.se$bill_length_mm)
ci.up<-(sps.avg$bill_length_mm)+1.96*(sps.se$bill_length_mm)

#par(family="Times New Roman") #Make the text pretty
bp<-barplot(as.matrix(t(sps.sex.avg)), #need to transpose table
            ylim=c(0,50), #set for your dataset
            beside=TRUE,
            las=1, #changes angle of xlab text, = 2 for vertical
            ylab="Bill length (mm)", #change for your label, always include units
            names.arg=c(as.character(sps.avg$species[1:3])), #names to use for bars
            col=c("light blue","royal blue")) #bar colors
legend("topleft", 
  legend = c("Female", "Male"), 
  col = c("light blue","royal blue"),
  pch = c(15,15), 
  bty = "n", 
  cex = 1.2, 
  text.col = "black")
arrows(x0=bp,x1=bp,y0=ci.low[c(1,4,2,5,3,6)],y1=ci.up[c(1,4,2,5,3,6)],code = 3, angle = 90, len = 0.02, xpd = NA)

#See how annoying that is? Let's do it in ggplot --much easier for multiple levels and groupings.
```

### Same plot, but in ggplot and using standard error bars
```{r}
spsex.avg<-aggregate(bill_length_mm~species+sex,data=penguins,FUN=mean) #get means for each species & sex

spsex.se<-aggregate(bill_length_mm~species+sex,data=penguins,FUN=my.se) #get standard error

#pull into one dataframe
bills.spsex<-cbind(spsex.avg,spsex.se$bill_length_mm)
colnames(bills.spsex)<-c("sps","sex","mean","se")

#plot
ggplot(bills.spsex,aes(x=sps,y=mean,fill=sex))+
  geom_bar(stat="identity",position=position_dodge(.9))+
  geom_errorbar(aes(ymax=mean + se, ymin=mean-se),width=0.2,position=position_dodge(.9))+
  scale_fill_manual(values=c("female" = "light blue","male"="royal blue"), drop = FALSE)+
  scale_x_discrete(limits=levels(bills.spsex$sps))+
  scale_y_continuous(breaks=seq(0,60,5),expand = c(0, 0),lim=c(0,60))+
  labs(x="",y="Bill Length (mm)")+
  theme(legend.position = "bottom",legend.text=element_text(size=12),legend.title=element_blank()) + #extra to make it pretty
  theme(panel.background = element_blank())+
  #theme(text = element_text(family = "Times New Roman"))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

### Relative abundance barplot if relative abundance was calculated in excel
```{r}
df<-read.csv("relabunddata.csv")
df
```


