---
title: "Understandng gglot"
output: html_notebook
---
In this notebook, we are going to iteratively walk through 'building' a graph using ggplot.
```{r}
library(ggplot2)
library(palmerpenguins)
```
Quick look at the data to figure out what we want to visualize:
```{r}
head(penguins)
summary(penguins)
penguins<-na.omit(penguins) #remove NAs
```
Let's visualize the relationship between bill length and flipper length.

First use R base plot:
```{r}
plot(penguins$bill_length_mm,penguins$flipper_length_mm)
```
This isn't pretty and we have to add a lot of code to make it pretty.

Now, let's use ggplot
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm)) #sets up the graph characteristics
```

```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm)) +
  geom_point() #adds the points

```

Let's fix the labels
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm)) +
  geom_point() +
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length") #add labels

```
Let's change the scales
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm)) +
  geom_point() +
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length") +
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+ #set y scale
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) #set x scale

```
Let's change the theme to get rid of the grey background
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm)) +
  geom_point() +
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length") +
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) +
  theme_bw() #changes the base style of the plot, see also theme_classic

```


Let's change the point characteristics
- http://www.sthda.com/english/wiki/ggplot2-point-shapes
- http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm)) +
  geom_point(shape=22,color="blue",fill="cyan",size=2.5) + #this will make all the points have the same characteristics
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length") +
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) +
  theme_bw() 
```
Now, let's change the point characteristics based on data attributes
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm,color=species, shape=sex)) + #add color and shape within the aes()
  geom_point(size=2) + 
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length") +
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) +
  theme_bw() 
```

Let's pick our own colors for species
- Another helpful tool for picking colors: https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm,color=species, shape=sex)) + 
  geom_point(size=2) + 
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length") +
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) +
  theme_bw() +
  scale_color_manual(values=c("Adelie"="#1b9e77","Chinstrap"="#d95f02","Gentoo"="#7570b3")) #set your own colors
```

Let's pick our own shapes
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm,color=species, shape=sex)) + 
  geom_point(size=2) + 
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length") +
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) +
  theme_bw() +
  scale_color_manual(values=c("Adelie"="#1b9e77","Chinstrap"="#d95f02","Gentoo"="#7570b3")) +
  scale_shape_manual(values=c("female"=17,"male"=16)) #pick your own shapes
```

Customize the legend
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm,color=species, shape=sex)) + 
  geom_point(size=2) + 
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length", color="Penguin Species", shape="Sex")+ #change legend labels 
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) +
  theme_bw() +
  scale_color_manual(values=c("Adelie"="#1b9e77","Chinstrap"="#d95f02","Gentoo"="#7570b3")) +
  scale_shape_manual(values=c("female"=17,"male"=16)) +
  theme(legend.position="bottom",legend.text = element_text(size=10),legend.title = element_text(size=12)) #move to bottom and change size of the text
  
```

Split plot into multiple panels by year
```{r}
ggplot(penguins,aes(x=bill_length_mm,y=flipper_length_mm,color=species, shape=sex)) + 
  geom_point(size=2) + 
  labs(x="Bill length (mm)",y="Flipper length (mm)", title="Relationship between Bill & Flipper Length", color="Penguin Species",shape="Sex")+ 
  scale_y_continuous(limits=c(150,250),breaks=c(150,175,200,225,250))+
  scale_x_continuous(limits=c(30,60),breaks=c(30,40,50,60)) +
  theme_bw() +
  scale_color_manual(values=c("Adelie"="#1b9e77","Chinstrap"="#d95f02","Gentoo"="#7570b3")) +
  scale_shape_manual(values=c("female"=17,"male"=16)) +
  theme(legend.position="bottom",legend.text = element_text(size=10),legend.title = element_text(size=12)) +
  facet_wrap(~year) #split data into multiple panels by year
  
```