---
title: "An Short Introduction to General Data Exploration "
output: html_notebook
---

Code written by Sara Williams
Contact: swilliams@mote.org


This is just a quick overview of some data exploration that I might use when first looking at a new dataset. First, I go through using base-R functions and then I use tidyverse functions.

### Packages used:

```{r}
library(palmerpenguins)
library(tidyverse)
```

Open-source data used in this notebook is from:
https://allisonhorst.github.io/palmerpenguins/index.html

```{r}
citation("palmerpenguins")
```

### What is the data?

One way to see this is to just print the dataframe:
```{r}
penguins
```

Better ways to look at large datasets:
```{r}
class(penguins)
#So it's a table dataframe

#Get a summary
summary(penguins)

#inspect just one column
class(penguins$species)

#get the structure which tells us about the variables
str(penguins)

#look at the first 6 rows:
head(penguins)

colnames(penguins)

dim(penguins)

nrow(penguins)
ncol(penguins)

length(penguins)
length(penguins$species)
length(levels(penguins$species))
```

### Subsetting the data

```{r}
#remember: [rows , columns]
#select just the first 3 columns, but all rows
penguins[,1:3]
#select row 47 and columns 1,4
penguins[47,c(1,4)]

#subset(): select particular columns from the data, reorder columns, and select particular rows from the data
subset(penguins,select=c("species")) #select one column
subset(penguins,select=c("island","species")) #reorder columns
subset(penguins,subset=species=="Adelie") #just one species
subset(penguins,subset=species!="Adelie",select=c("species","island")) #all species not Adelie
subset(penguins,subset=bill_length_mm>40) #observations with bill lenght > 40 mm
```


### Dealing with missing data
```{r}

is.na(penguins$bill_depth_mm) #logical vector of bill depths that are NA
penguins$bill_depth_mm[!(is.na(penguins$bill_depth_mm))] #! gives all of the non NA values
na.omit(penguins$bill_depth_mm) #changes it to an omit class, will tell you which indexes were omitted
penguins$bill_depth_mm[complete.cases(penguins$bill_depth_mm)] #another way of getting values that arent NA

```

### Refresher on functions to summarize/describe quantitative data
```{r}

mean(penguins$bill_depth_mm) #using the base mean function gets an error because NA's
mean(penguins$bill_depth_mm,na.rm=TRUE) #excludes NA

median(penguins$bill_depth_mm,na.rm=TRUE)

mysd<-sd(penguins$bill_depth_mm,na.rm=TRUE)

mysd/sqrt(length(penguins$bill_depth_mm)) #standard error

#not useful here but...
rowSums(penguins[3:5])
colSums(penguins[3:5],na.rm=TRUE)
```


### Now what about qualitative data like factors?
```{r}
class(penguins$species)
levels(penguins$species) #what are the values (levels) of the factor species

penguins$sex<-factor(penguins$sex,levels=c("male","female")) #will reorder the levels, but can't just use diffent words, have to stick to originial level names
# renaming factors ####
levels(penguins$sex)
levels(penguins$sex)[1]<-"m"
levels(penguins$sex)[2]<-"f"

#using table 
table(penguins$species,penguins$island)
table(penguins$species,penguins$sex)
```



### Using the aggregate function
When at least one variable is categorical (a factor), we may be interested in a summary statistic
for each category. An common task when dealing with large and complex datasets it to compute
summary statistics such as the mean or the standard deviation for subsets of the data. This can
be accomplished by using function aggregate (type ?aggregate at the console for help).
The first argument of aggregate is called a formula. Names that appear on the left of the ⇠
correspond response variables and names that appear to the right of the ⇠ correspond to explanatory
variables. In this case, the code states that we want to apply some function to weight
based on the distinct values of feed. The second argument FUN=mean specifies that the FUNction in
question is the mean, and the last argument data=chickwts specifies the variable in which weight
and feed are stored.
```{r}
#get mean of bill length by species
aggregate(bill_length_mm~species,FUN=mean,data=penguins)
#now by species and island combo
aggregate(bill_length_mm~species+island,FUN=mean,data=penguins)
```

### Some basic plots to explore the data
```{r}
#scatterplot - numeric & numeric
plot(penguins$flipper_length_mm,penguins$body_mass_g)
```

```{r}
#histogram - distribution of one numeric catergory
hist(penguins$flipper_length_mm)
range(penguins$flipper_length_mm,na.rm=TRUE) #range will tell you max and min

# now, 3 panels, one for each species
par(mfrow=c(3,1))
hist(subset(penguins,subset=species=="Adelie")$flipper_length_mm,main="Adelie Flipper Length",xlim=c(170,240))
hist(subset(penguins,subset=species=="Gentoo")$flipper_length_mm,main="Gentoo Flipper Length",xlim=c(170,240))
hist(subset(penguins,subset=species=="Chinstrap")$flipper_length_mm,main="Chinstrap Flipper Length",xlim=c(170,240))
```

```{r}
#boxplot: visualize more info about categorical and numeric data
plot(penguins$bill_length_mm~penguins$species)
```

## Now same sorts of things, but using the tidyverse
```{r}
glimpse(penguins)
```
### Using pipes
```{r}
# Piping ####
penguins %>%
  select(where(is.factor)) %>%
  glimpse()

#count penguins by species and island ... similiar to table function
penguins %>%
  count(species, island, .drop = FALSE)

# Count penguins for each species / sex
penguins %>%
  count(species, sex, .drop = FALSE)

penguins %>%
  select(body_mass_g, ends_with("_mm")) %>%
  glimpse()

#will filter to select rows with weight less than 3000 and choose to keep only 3 columns
penguins_w<- penguins %>%
  filter(body_mass_g < 3000) %>% 
  select(species, sex, body_mass_g)

#challenge, subset data to include individuals before 1995, retain columns year, sex, weight
penguins09<- penguins %>%
  filter(year == 2009) %>% 
  select(year, sex, species)

# add columns with mutate() ####
penguins2<-penguins %>%
  mutate(body_mass_kg=body_mass_g/1000) %>%
  head

penguins %>%
  filter(!is.na(body_mass_g)) %>%
  mutate(body_mass_kg=body_mass_g/1000) %>%
  head

#summarizing with split-apply-combine
penguins%>%
  group_by(sex) %>%
  summarize(mean_body_mass_g = mean(body_mass_g, na.rm = TRUE))

```

```{r}
# Scatterplot example 1: penguin flipper length versus body mass
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species,shape = species),size = 2) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4"))
```
```{r}
# Histogram example: flipper length by species
ggplot(data = penguins, aes(x = flipper_length_mm)) +
  geom_histogram(aes(fill = species), alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4"))
```
```{r}
ggplot(penguins,aes(x=species,y=flipper_length_mm))+
  geom_boxplot()
```







